---
title: "Non-parametric survival methods in R: single group"
author: "Shannon Pileggi"
date: "February 9, 2018"
output: 
  html_document:
    highlight: tango
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=5, fig.height=3) 
```

# Getting started

1. Install the `survival` package.  This can be done through `Packages` -> `Install` or with

```{r, eval = F}
install.packages("survival")
```

This should only need to be done once on your personal computer; you may or may not need to do this for each session on the library computers.

2. Load the `survival` package.


```{r, eval = T}
library(survival)
```

# Basic survival

## Create a survival object

A survival object consists of two vectors:

 + time to event

 + censoring status
 
For example, suppose we have the following 7 motorist reaction times:
 
  `1.41 1.41+ 2.76+ 3.56 4.18 4.71+ 13.18`
  
From this, we can construct a vector of times and a vector with the censoring status (1 = complete, 0 = right-censored).  In this case, I will create the vectors in  a data frame.

**Note:**  You do **not** need to do this `data.frame` step if you have imported the data as a data frame with a command like `read.csv` or `read.table`.  This was only used to construct a data set from "raw" data.

```{r}
motorists <- data.frame(time = c(1.41, 1.41, 2.76, 3.56, 4.18, 4.71, 13.18),
                        censor = c(1, 0, 0, 1, 1, 0, 1))
motorists
```


A survival object uses the `Surv` function to combine the `time` and `censor` vectors.

```{r}
motorists_surv <- Surv(motorists$time, motorists$censor)
```

## Obtain Kaplan-Meier estimates

Use the `survfit` function on your survival object to obtain Kaplan-Meier estimates.  Here, we are using modeling syntax where `~ 1` represents that we are creating an intercept only model.
```{r}
KM_obj <- survfit(motorists_surv ~ 1, 
                  conf.type = "none")
```

Then produce a summary of your Kaplan-Meier object with the `summary` function.
```{r}
summary(KM_obj)
```

You can also get additional information from your Kaplan-Meier object just by submitting the name of the object - this give the sample size, the number of events, and the estimated median survival time.
```{r}
KM_obj
```

## Create a survival object when you obtain KM estimates

You don't have to first specify the survival object and then use `survfit` - this can be done in one step as:

```{r}
KM_obj <- survfit(Surv(time, censor) ~ 1, 
                  data = motorists, 
                  conf.type = "none")

summary(KM_obj)
```

## Plotting

### base R

To plot the Kaplan-Meier curve, just execute the `plot` function on the `KM_obj`.

```{r}
plot(KM_obj,
     xlab = "Seconds", 
     ylab = "Survival Probability", 
     main = "KM Curve with Confidence Limits")
```


### ggsurvplot

To use a `gg` version of survival plotting, install the `survminer` package and use the `ggsurvplot` function.  Note that this function has many options, including whether or not to plot confidence intervals (`conf.int`) or the risk table (`risk.table`).  Note that I received an error in `ggsurvplot` when I had `conf.type = "none"` in my `KM_obj`, but when I set `conf.type = "plain"` I no longer received an error.

```{r, message = FALSE}
library(survminer)
```

```{r, fig.height=5}
KM_obj <- survfit(Surv(time, censor) ~ 1, 
                  data = motorists, 
                  conf.type = "plain")

ggsurvplot(KM_obj, data = motorists, conf.int = FALSE, risk.table = TRUE)
```

## Confidence limits

For `conf.type`:

 + `none` omits confidence intervals

 + `plain` computes standard confidence intervals
 
 + `log-log` computes confidence intervals based on the log-log transformation
  
  
Plotting a Kaplan-Meier object that specified confidence limits will automatically plot the confidence limits.

```{r}
KM_obj <- survfit(Surv(time, censor) ~ 1, 
                  data = motorists, 
                  conf.type = "plain")

summary(KM_obj)

plot(KM_obj,
     xlab = "Seconds", 
     ylab = "Survival Probability", 
     main = "KM Curve with Confidence Limits")
```

## Nelson-Aelen Estimator of $S(t)$

Use the option `type = "fh"` in the `survfit` function to compute Nelson-Aalen estimates of survival probabilities.
```{r, include = F}
source("C:/Users/spileggi/Google Drive/STAT 417/plot.haz.R")
```


```{r}
KM_obj_NA <- survfit(Surv(time, censor) ~ 1, 
                     data = motorists, 
                     type = "fh",
                     conf.type = "none")

summary(KM_obj_NA)

plot(KM_obj_NA,
     xlab = "Seconds", 
     ylab = "Survival Probability", 
     main = "S(t) estimated by Nelson-Aelen")
```

# Hazard / Cumulative Hazard

## Overview

There are not any built in functions in the `survival` package that plot the estimated hazard rates or cumulative hazards.  Two supplemental `R` functions have been written to plot the the Kaplan-Meier estimated hazard function and the Nelson-Aalen cumulative hazard function.  The `plot.haz.R` script in the PolyLearn `Data sets` folder contains two functions:

+ `plot_haz` plots the Kaplan-Meier estimated hazard function $\hat{h}(t)$

+ `plot_chaz` plots the Nelson-Aalen estimated cumulative hazard function $\tilde{H}(t)$

## Accessing the functions

1.  Download the R script `plot.haz.R` from PolyLearn and open it in RStudio.

2.  With this file open, click on `Source`.

3.  Now you should be able to see the two functions `plot_haz` and `plot_chaz` in your `Environment`.

4.  To include the source code in an R script or Rmarkdown file:

+ Click on `History` in the upper right hand corner of RStudio.

+ Scroll down until you see the line of code that begins with `source`.  Click on ths line of code.  For example:

```{r, eval=F}
source("C:/Users/spileggi/Google Drive/STAT 417/plot.haz.R")
```

+ Once that line of code is highlighted, click on `To source` and the R code should now appear in your upper left-hand panel.  If you are working in an Rmarkdown file (.Rmd), be sure to place this code in a code chunk.

## Plotting

Both functions can be applied to an object created by `survifit`.


```{r}
KM_obj <- survfit(Surv(time, censor) ~ 1, 
                  data = motorists, 
                  conf.type = "none")

plot_haz(KM_obj)

plot_chaz(KM_obj)

```

