---
title: "Non-parametric survival methods in R: comparing groups"
author: "Shannon Pileggi"
date: "February 9, 2018"
output: 
  html_document:
    highlight: tango
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

# Getting started

For illustration, we'll use the `ovarian` data set located in the `survival` library.  Recall that this data compares two treatments for ovarian cancer.

+ `futime` = time to death

+ `fustat` = censoring status (1 = complete, 0 = right censored)

+ `rx` =  treatment group (1/2)

To access the data set, execute the code below.

```{r}
# ovarian data set is in the survival library
library(survival)
data(ovarian) 
```

Next, create your `survfit` model.  In previous models, we used the syntax ` ~ 1` to indicate an intercept only model (not stratified).  Now, we'll use the syntax ` ~ group_var` to indicate to compute Kaplan-Meier curves separately for each group.  In this case, `rx` defines our groups of interest.

```{r}
KM_obj <- survfit(Surv(futime, fustat) ~ rx, data = ovarian)
```

**Note:**  If your `group_var` variable is coded numerically (as in 1/2, etc.) you need to explicitly tell R to treat this as a categorical rather than quantitative variable.  You do this by wrapping the `group_var` variable in the `as.factor` function. A generic expression for this would look like:

```{r, eval = FALSE}
KM_obj <- survfit(Surv(time, censor) ~ as.factor(group_var), data = my_data)
```

# Obtaining estimates

Obtain the Kaplan-Meier survival estimates by group (`rx` in this case) by using the `summary` function:

```{r}
summary(KM_obj)
```

You can also obtain esimates of the median survival time in each group simply by printing the Kaplan-Meier object.

```{r}
KM_obj
```


# Plots

## base R

When plotting in base R, you need to specify a graphical element to differentiate the curves.  You can do this by color (`col`) or line type (`lty`) as shown belown.  In this case, we have two groups, so I specified two different line types.  Similarly, the two graphical elements need to be defined in the legend as well.

```{r, fig.height=4}
plot(KM_obj, 
     lty = 1:2,
     xlab = "Time to death",
     ylab = "Survival probability",
     main = "KM curves for two ovarian cancer treatments"
     )

legend("topright", c("Trt 1", "Trt 2"), lty = 1:2)
```


## ggsurvplot

To use a `gg` version of survival plotting, install the `survminer` package and use the `ggsurvplot` function.  Note that this function has many options, including whether or not to plot confidence intervals (`conf.int`) or the risk table (`risk.table`).

```{r, message = FALSE}
library(survminer)
```


```{r, fig.height=5}
ggsurvplot(KM_obj, data = ovarian, risk.table = TRUE)
```

# Log-rank test

Use the function `survdiff` on a survival object to perform the log-rank test.

```{r}
survdiff(Surv(futime, fustat) ~ rx, data = ovarian)
```

# Notes

+ `R` does not provide the Wilcoxon test.

+ Methods are the same if you have more than 2 groups.

