---
title: "Lab 6"
author: "Your name"
date: "February 27, 2018"
output:
  pdf_document:
    toc: yes
  html_document:
    highlight: tango
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---


```{r}
library(survival)
```

# The data

Recall the `Rearrest.txt` data which contains the following variables:

Variable  | Meaning
--------- |-----------------------------------------------
`months`  | time (in months) from prison release to re-arrest
`censor`  | the censoring status indicator variable
`personal`| `personal=1` if former inmate had a history of person-related crimes  (i.e. those with one or more convictions for offenses such as aggravated assault or kidnapping), and 0 otherwise.
`property`|  `property=1` if former inmate was convicted of a property-related crime, 0 otherwise.
`cenage`  | the "centered"" age of individual, i.e. the difference between the age of the individual upon release and the average age of all inmates in the study. For example, a subject with value 8.3 for `cenage` is at the time of release, 8.3 years older than the average age of all former inmates.

To import the data set, use `read.table`.

```{r, eval = F}
rearrest <- read.table("FILE PATH/Rearrest.txt", header = TRUE)
```



# Question 1

## Question 1a
*Consider fiting a Cox regression model to the `rearrest` data with one categorical predictor of `property` (you don't need to do it yet).  Should we use `as.factor()` on the `property` variable?  Why or why not?*



## Question 1b
*Fit the Cox regression model to the `rearrest` data with one categorical predictor of `property`.  Report and interpret the parameter estimate $\hat{\beta}$ in the context of the problem.*


## Question 1c
*Find and interpret (in context) the hazard ratio comparing former inmates who did and did not commit property related crimes.*



## Question 1d
*Report the 95% confidence interval for $\beta$ and interpret it in conxtext.*



## Question 1e
*Report the 95% confidence interval for the $HR$ and interpret it in conxtext.*



## Question 1f
*Conduct the Wald test at the $\alpha=0.05$ level of significance to dtermine if the risk of re-arrest differs for former inmates who did and did not commit property related crimes.  Be sure to report the test statistic and the $p$-value, and state a conclusion in context.*



## Question 1g 
*If we assume that the "proportional hazards" condition is satisfied, briefly explain what that means in the context of this problem.*







# Question 2

## Question 2a 
*Fit a one predictor Cox regression model to the \texttt{rearrest} data with the quantitative predictor `cenage`. Report and interpret the parameter estimate $\hat{\beta}$ in the context of the problem.*


## Question 2b 
*Find and **carefully** interpret (in context) the hazard ratio corresponding to a one year increase in `cenage`.*


## Question 2c 
*Compute and interpret the hazard ratio corresponding to a 5 year increase in `cenage`.*


## Question 2d
*Conduct the Wald test at the $\alpha=0.05$ level of significance to dtermine if the risk of re-arrest is associated with `cenage`.  Be sure to report the test statistic and the $p$-value, and state a conclusion in context.*

